---
title: "Chapter 4 Exercise Questions"
author: "Usman Khaliq"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(rethinking)
data("Howell1")

#===============================================================================

```

4E1. In the model definition below, which line is the likelihood?

yi~Normal(μ,σ)
μ~Normal(0,10)
σ~Uniform(0,10)

Answer: The first line is the likelihood.

4E2. In the model definition just above, how many parameters are in the posterior distribution? 

Answer: There are two parameters in the posterior distribution. 

4E3. Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.

Answer: 
$$\Pr(\mu,\sigma|y) = \frac{\prod_i \mathrm{Normal} (y_i|\mu,\sigma) \mathrm{Normal} (\mu|0,10) \mathrm{Uniform}(\sigma|0,10)}{\int \int \prod_i \mathrm{Normal}(h_i|\mu,\sigma) \mathrm{Normal}(\mu|0,10) \mathrm{Uniform}(\sigma|0,10)d\mu d\sigma}$$


4E4. In the model definition below, which line is the linear model?

yi~Normal(μ,σ)
μi=α+βxiα~Normal(0,10)
β~Normal(0,1)
σ~Uniform(0,10)

Answer: The second line is the linear model. 

4E5. In the model definition just above, how many parameters are in the posterior distribution?

Answer: There are three parameters in the posterior distribution. 

4M1. For the model definition below, simulate observed heights from the prior (not the posterior).

yi~Normal(μ,σ)
μ~Normal(0,10)
σ~Uniform(0,10) 

```{r}

sample_mu <- rnorm(1e4, 0, 10)
sample_sigma <- runif(1e4, 0, 10)
prior_y <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_y)

```

4M2. Translate the model just above into a map formula.

```{r, eval=FALSE}
model <-alist(
    y ~ dnorm(mu, sigma),
    mu ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  )

```

4M3. Translate the map model formula below into a mathematical model definition.

flist <- alist(

y ~ dnorm( mu , sigma ),

mu <- a + b*x,

a ~ dnorm( 0 , 50 ),

b ~ dunif( 0, 10
sigma ~ dunif( 0, 50)

)

$$
\begin{aligned}
y_i &\sim \mathrm{Normal}(\mu, \sigma) \\
\mu_i &= \alpha + \beta x_i \\
\alpha &\sim \mathrm{Normal}(0, 50) \\
\beta &\sim \mathrm{Uniform}(0, 10) \\
\sigma &\sim \mathrm{Uniform}(0, 50)
\end{aligned}
$$


4M4. A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

Answer.
My mathematical model will be as follows:
yi~Normal(μi,σ)
μi=α+β * age
α ~ Normal(120, 10)
β~Normal(4,2)
σ~Uniform(0,50) 

4M5. Now suppose I tell you that the average height in the first year was 120 cm and that every student got taller each year. Does this information lead you to change your choice of priors? How?

This information just makes me more confident about the average height of the students that I chose. 

4M6. Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm. How does this lead you to revise your priors?

I would adjust the prior of σ as follows:
σ ~ Uniform(0, 8) 

4H1 The weights listed below were recorded in the !Kung census, but heights were not recorded for these individuals. Provide predicted heights and 89% intervals (either HPDI or PI) for each of these individuals. That is, fill in the table below, using model-based predictions.


> Individual | weight | expected height | 89% interval
> ---------- | ------ | --------------- | ------------
> 1 | 46.95 | | 
> 2 | 43.72 | |
> 3 | 64.78 | |
> 4 | 32.59 | |
> 5 | 54.63 | |

```{r}
d <- Howell1
model <- rethinking::map(
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ a + b * weight,
    a ~ dnorm(140, 30),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = d
) 

values <- c(46.95, 43.72, 64.78, 32.59, 54.63)
predicted_heights <- link(model, data = data.frame(weight = values))
```

```{r}
predicted_heights_means <- apply(predicted_heights, 2, mean)
predicted_hpdi <- apply(predicted_heights, 2, HPDI, prob = 0.89)
```

```{r}
data.frame(
  individual = 1:5,
  weight = values,
  expected_height = predicted_heights_means,
  lower_interval = predicted_hpdi[1,],
  upper_interval = predicted_hpdi[2,]
)
```

4H2. Select out all the rows in the Howelll data with ages below 18 years of age. If you do it right, you should end up with a new data frame with 192 rows in it.

```{r}
filtered_data <- 
  d %>% 
  filter(age < 18)
nrow(filtered_data)
```

(a)  Fit a linear regression to these data, using map. Present and interpret the estimates. For every 10 units of increase in weight, how much taller does the model predict a child gets?

```{r}
model <- rethinking::map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b * weight,
    a ~ dnorm(110, 30),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0, 60)
  ),
  data = filtered_data
)
```

```{r}
precis(model, corr = TRUE)
```

For every 10 unit increase in weight, the model assumes an increase of height of 27.2 cm. 

(b)  Plot the raw data, with height on the vertical axis and weight on the horizontal axis. Superimpose the MAP regression line and 89% HPDI for the mean. Also superimpose the 89% HPDI for predicted heights.

```{r}
weight.seq <- 
  seq(
    from = min(filtered_data$weight),
    to = max(filtered_data$weight),
    by = 1
  )

mu <- link(model, data = data.frame(weight = weight.seq))
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.89) 
sim.height <- sim(model, data = list(weight = weight.seq))
height.HPDI <- apply(sim.height, 2, HPDI, prob = 0.89)
```

```{r}
plot(height ~ weight, data = filtered_data, col = col.alpha(rangi2, 0.5))

lines(weight.seq, mu.mean)

shade(mu.HPDI, weight.seq) 

shade(height.HPDI, weight.seq)
```

(c)  What aspects of the model fit concern you? Describe the kinds of assumptions you would change, if any, to improve the model. You don’t have to write any new code. Just explain what the model appears to be doing a bad job of, and what you hypothesize would be a better model.

The linear model is doing a poor job of fitting the data, both at higher and lower values of weights. It seems like the linear model is a poor fit for the data. 

4H3. Suppose a colleague of yours, who works on allometry, glances at the practice problems just above. Your colleague exclaims, “That’s silly. Everyone knows that it’s only the logarithm of body weight that scales with height!” Let’s take your colleague’s advice and see what happens.

(a)  Model the relationship between height (cm) and the natural logarithm of weight (log-kg). Use the entire Howelll data frame, all 544 rows, adults and non-adults. Fit this model, using quadratic approximation:

hi~Normal(μi,σ)

μi=α+βlog(wi)

α~Normal(178,100)

β~Normal(0,100)

σ~Uniform(0,50)

where hi is the height of individual i and wi is the weight (in kg) of individual i. The function for computing a natural log in R is just log. Can you interpret the resulting estimates?

```{r}
model <- rethinking::map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b * log(weight),
    a ~ dnorm(178, 100),
    b ~ dnorm(0, 100),
    sigma ~ dunif(0, 50)
  ),
  data = d
)
```

Then use samples from the quadratic approximate posterior of the model in (a) to superimpose on the plot: (1) the predicted mean height as a function of weight, (2) the 97% HPDI for the mean, and (3) the 97% HPDI for predicted heights. 

```{r}
weight.seq <- 
  seq(
    from = min(d$weight),
    to = max(d$weight),
    by = 1
  )

mu <- link(model, data = data.frame(weight = weight.seq))
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.97) 
sim.height <- sim(model, data = list(weight = weight.seq))
height.HPDI <- apply(sim.height, 2, HPDI, prob = 0.97)

```

```{r}
plot(height ~ weight, data = d, col = col.alpha(rangi2, 0.5))

lines(weight.seq, mu.mean)

shade(mu.HPDI, weight.seq) 

shade(height.HPDI, weight.seq)

```

