---
title: "Chapter 7 - Interactions"
author: "Usman Khaliq"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(rethinking)
# Parameters

#===============================================================================

# Code
```


R 7.1 
Lets explore regressions of terrain ruggedness against economic performance in Africa and outside of Africa

```{r}
data("rugged")
d <- rugged
#transform outcome to log scale
d$log_gdp <- log(d$rgdppc_2000)

#remove countries which do not have GDP data
dd <- 
  d %>%
  drop_na(rgdppc_2000)

#split countries on whether they are in africa or not in africa
d.A1 <- 
  dd  %>% 
  filter(cont_africa == 1)

d.A0 <- 
  dd %>% 
  filter(cont_africa == 0)
  
```

R Code 7.2

Fit regression models to the data above.

```{r}
#African nations
m7.1 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d.A1
) 

#non-African nations
m7.2 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d.A0
) 
```

```{r}
#define sequence of rugged terrain values to make predictions on
rugged_seq_africa <- 
  seq(from = min(d.A1$rugged), to = max(d.A1$rugged), length.out = 30) 

mu_africa <- link(m7.1, data = data.frame(rugged = rugged_seq_africa))
mu.mean <- apply(mu_africa, 2, mean)
mu.PI <- apply(mu_africa, 2, HPDI, prob = 0.89) 


plot(log_gdp ~ rugged, data = d.A1, col = col.alpha(rangi2, 0.5))
mtext("Africa")
#draw MAP line
lines(rugged_seq_africa, mu.mean)

#draw HPDI region for line
shade(mu.PI, rugged_seq_africa)


```


```{r}
#define sequence of rugged terrain values to make predictions on
rugged_seq_not_africa <- 
  seq(from = min(d.A0$rugged), to = max(d.A0$rugged), length.out = 30) 

mu_not_africa <- link(m7.2, data = data.frame(rugged = rugged_seq_not_africa))
mu.mean <- apply(mu_not_africa, 2, mean)
mu.PI <- apply(mu_not_africa, 2, HPDI, prob = 0.89) 


plot(log_gdp ~ rugged, data = d.A0, col = col.alpha(rangi2, 0.5))
mtext("not Africa")
#draw MAP line
lines(rugged_seq_not_africa, mu.mean)

#draw HPDI region for line
shade(mu.PI, rugged_seq_not_africa)

```

Its not a good idea to split data, like we split the data above. The following are the reasons for it:

1) Some parameters, such as sigma, were assumed to not depend on a country's Africa identity. By splitting the data, we are hurting the accuracy of such estimates because we are now making two less accurate estimates instead of pooling all data into making one solid estimate. 

2) In order to make probability estimates about the `cont_africa` variable, we need to include it in the data that the model works on. Unless our model analyses all the data, it canot make a solid estimate of the uncertainty in the probability that `cont_africa` has on the outcome variable.

3) We might need to use information criteria to compare how different models treat different continents. In order for the estimate to be as accurate as possible, the models should read all the data.

4) In multilevel models, we can in fact learn across different categories. Therefore, it makes sense to not split data before inputting it into the model. 

Lets see if adding the categorical variable(dummy variable) `cont_africa` will reveal the reverse slope in the model.

R Code 7.3

The following is a simple linear regression model for `log_gdp` on ruggedness over the entire dataset  

```{r}
m7.3 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)
```

The second is a model that includes a dummy variable for African countries.

R Code 7.4

```{r}
m7.4 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged + bA * cont_africa,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)

```

Now, lets compare the WAIC values of these models

R Code 7.5

```{r}
compare(m7.3, m7.4)
```

From the above, we can see that all weight is assigned to model m7.4. The standard error in the difference of WAIC between the two models is 15, but the difference in WAIC values is 63, which implies a 95% confidence interval difference of 63 +- 30. Thus, we can gauge from these numbers that the dummy variable is picking up some important information since m7.4 is superior to the simple model. 

Now, lets plot the posterior prediction of m7.4 to see whether it achieves different slopes within and outside of Africa. 

R Code 7.6

```{r}
rugged.seq <- seq(from = -1, to = 8, by = 0.25)

#compute mu over samples, fixing cont_africa = 0
mu.NotAfrica <- 
  link(m7.4, data = data.frame(cont_africa = 0, rugged = rugged.seq))

#compute mu over samples, fixing cont_africa = 1
mu.Africa <- 
  link(m7.4, data = data.frame(cont_africa = 1, rugged = rugged.seq))

#summarise to means and intervals 
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob = 0.97)
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob = 0.97)

```

```{r}
plot(log_gdp ~ rugged, data = dd) 

#plot the regression lines
lines(rugged.seq, mu.NotAfrica.mean)
lines(rugged.seq, mu.Africa.mean, col = col.alpha(rangi2, 0.5))

#draw HPDI region for lines
shade(mu.NotAfrica.PI, rugged.seq)
shade(mu.Africa.PI, rugged.seq, col = col.alpha(rangi2, 0.5))
mtext("Africa is in Blue, Other Countries in Black")


```

From the above, we can see that by just adding a dummy variable to the model, the model still predicted a poor relationship between ruggedness and a country's GDP. What the model only did was that it predicted a lower average GDP values for African countries, which can be shown from the fact that the regression line for Africa is lower. 

Now, instead, lets add a linear model that models the linear interaction effect between ruggedness and African nations. In short, the relationship between GDP and ruggedness will be conditional on whether a country is in Africa or not. 

Yi ~ Normal(µi, σ) [likelihood] 
µi = α + γiRi + βAAi [linear model of µ]
γi = βR + βARAi [linear model of slope] 

The third line above is the linear interaction effect between ruggedness and whether a country is in Africa or not. 

Lets model the above linear interaction.

R Code 7.7

```{r}
m7.5 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + gamma * rugged + bA * cont_africa,
    gamma <- bR + bAR * cont_africa,
    a ~ dnorm(8, 100),
    bA ~ dnorm(0, 1),
    bR ~ dnorm(0, 1),
    bAR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)
```

Now, lets compare this model using WAIC values with the previous models.

```{r}
compare(m7.3, m7.4, m7.5)
```

Model 7.5 has a weight of 0.97. However, the weight of 0.03 that is given to model m7.4 shows that there is slight overfitting happening in m7.5. Also, the difference in the standard errors between the top two models is almost the same. This phenonema might also be there because there are only a limited number of African countries, and hence there is some sparsity in the data. 

R Code 7.10
First, lets calculate the posterior mean line and intervals for both plots. 

```{r}
rugged.seq <- seq(from = -1, to = 8, by = 0.25)

mu.Africa <- link(m7.5, data = data.frame(cont_africa = 1, rugged = rugged.seq))
mu.Africa.mean <- apply(mu.Africa$mu, 2, mean)
mu.Africa.PI <- apply(mu.Africa$mu, 2, PI, prob = 0.97)

mu.NotAfrica <- 
  link(m7.5, data = data.frame(cont_africa = 0, rugged = rugged.seq))
mu.NotAfrica.mean <- apply(mu.NotAfrica$mu, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica$mu, 2, PI, prob = 0.97)


```

now, lets plot the plots

R Code 7.11

```{r}
#plot African nations with regression
plot(
  log(rgdppc_2000) ~ rugged,
  data = d.A1,
  col = rangi2,
  ylab = "log GDP Year 2000",
  xlab = "Terrain Ruggedness Index"
  ) 
mtext("African Nations", 3)
lines(rugged.seq, mu.Africa.mean, col = rangi2)
shade(mu.Africa.PI, rugged.seq, col = col.alpha(rangi2, 0.3))

#plot non-African nations with regression
plot(
  log(rgdppc_2000) ~ rugged,
  data = d.A0,
  col = "black",
  ylab = "log GDP Year 2000",
  xlab = "Terrain Ruggedness Index"
  ) 
mtext("Non-African Nations", 3)
lines(rugged.seq, mu.NotAfrica.mean)
shade(mu.NotAfrica.PI, rugged.seq)

```

Now above, we can see that the slopes reverse direction inside and outside of Africa. 

Interpreting interaction estimates is tricky. The following are two reasons for being wary about directly interpreting tables of posterior means and standard deviations.

1) When we add interactions to a model, it changes the meaning of parameters. A "main effect" coefficient in the interaction model does not mean the same thing as a coefficient of the same name in a model without an interaction. In a simple linear regression, with no interactions, each coefficient tells us how much the average outcome changes when the predictor changes by one unit. However, in case of an interaction, this does not hold true anymore, since the coefficient is now dependent on other predictor variables. We can no longer read these values directly from tables. 

For instance, in the above example, the parameter estimates are as follows:

```{r}
precis(m7.5)
```

In the above table, we do not see values for gamma, since it wasn't estimated. However, we would have to estimate it ourselves for both within Africa and outside of Africa.

Within Africa, gamma is as follows

```{r}
-0.18 + 0.35
```

Outside of Africa, gamma is as follows:

```{r}
-0.18 + 0.35 * 0
```

2) It is difficult to incorporate uncertainty just by looking at tables of numbers. Since gamma depends on parameters,and those parameters have posterior distributions, gamma also has a posterior distribution. 

We can compute the posterior distribution of gamma as follows:

```{r}
post <- extract.samples(m7.5)
gamma.Africa <- post$bR + post$bAR * 1
gamma.notAfrica <- post$bR + post$bAR * 0
```

The means of these posterior distributions for gamma values is as follows:

```{r}
mean(gamma.Africa)
mean(gamma.notAfrica)
```

These means are very close to the MAP values of mean calculated above. 

Now, lets plot the distributions of gamma values within and outside of Africa to see where they overlap.

```{r}
dens(
  gamma.Africa, 
  xlim = c(-0.5, 0.6),
  ylim = c(0, 5.5),
  xlab = "gamma",
  col = rangi2
  )
dens(gamma.notAfrica, add = TRUE)
```

Linear interactions are symmetrical, and our golum cannot tell the difference between one symmetrical interaction and the other. Consider the following:

Yi ~ Normal (µi σ) [likelihood]
µi = α + γiRi + βAAi [linear model of µ]
γi = βR + βARAi [linear model of slope] 

Lets now expand γi into the expression for µi

μi=α+(βR+βARAi)Ri+βAAi=α+βRRi+βARAiRi+βAAi 

Now factor together the terms with Ai in them:

μi=α+βRRi+(βA+βARRi)Ai

From the above mathematical notations,we can see that for our linear interaction model, the following two phrasings are equally valid.

1) How much does the influence of ruggedness on GDP depend upon whether the nation is in Africa?

2) How much does the influence of being in Africa on GDP depend upon ruggedness? 

Lets plot the second interpretation below. 

R Code 7.17

```{r}
#get minimum and maximum rugged values
q.rugged <- range(dd$rugged)

#compute lines and confidence intervals
mu.ruggedlo <- 
  link(m7.5, data = data.frame(rugged = q.rugged[1], cont_africa = 0:1)) 

mu.ruggedlo.mean <- apply(mu.ruggedlo$mu, 2, mean)
mu.ruggedlo.PI <- apply(mu.ruggedlo$mu, 2, PI) 

mu.ruggedhi <- 
  link(m7.5, data = data.frame(rugged = q.rugged[2], cont_africa = 0:1)) 

mu.ruggedhi.mean <- apply(mu.ruggedhi$mu, 2, mean)
mu.ruggedhi.PI <- apply(mu.ruggedhi$mu, 2, PI) 

#plot it all, splitting points at  median
med.r <- median(dd$rugged)
ox <- ifelse(dd$rugged > med.r, 0.05, -0.05)
plot(
  dd$cont_africa + ox,
  log(dd$rgdppc_2000),
  col = ifelse(dd$rugged > med.r, rangi2, "black"),
  xlim = c(-0.25, 1.25),
  xaxt = "n",
  ylab = "log GDP year 2000",
  xlab = "continent"
  ) 
axis(1, at = c(0, 1), labels = c("other", "Africa"))
lines(0:1, mu.ruggedlo.mean, lty = 2, col = "black")
shade(mu.ruggedlo.PI, 0:1)
lines(0:1, mu.ruggedhi.mean, col = rangi2)
shade(mu.ruggedhi.PI, 0:1, col = col.alpha(rangi2, 0.25))


```

In the above, blue points are countries with terrain ruggedness above the median. Black points are countries with terrain ruggedness below the median. We can see from the plot that countries in Africa on avearge have lower GDP(due to colonization). The black line and black shaded region tell us the expected reduction in log GDP when we take a nation with minimum terrain ruggedness and change its continent. 

The blue line and blue shaded region show the change in GDP for a nation which has maximum terrain ruggedness between. We can see from this line that changing the continent has a very negligible effect on the GDP value. However, given the large area of the shaded region, this estimate is not very accurate. Thus, for a nation with veyr high rugged terrain, there is negligible effect on GDP by being in Africa. 

