---
title: "Chapter 7 - Interactions"
author: "Usman Khaliq"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(rethinking)
# Parameters

#===============================================================================

# Code
```


R 7.1 
Lets explore regressions of terrain ruggedness against economic performance in Africa and outside of Africa

```{r}
data("rugged")
d <- rugged
#transform outcome to log scale
d$log_gdp <- log(d$rgdppc_2000)

#remove countries which do not have GDP data
dd <- 
  d %>%
  drop_na(rgdppc_2000)

#split countries on whether they are in africa or not in africa
d.A1 <- 
  dd  %>% 
  filter(cont_africa == 1)

d.A0 <- 
  dd %>% 
  filter(cont_africa == 0)
  
```

R Code 7.2

Fit regression models to the data above.

```{r}
#African nations
m7.1 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d.A1
) 

#non-African nations
m7.2 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d.A0
) 
```

```{r}
#define sequence of rugged terrain values to make predictions on
rugged_seq_africa <- 
  seq(from = min(d.A1$rugged), to = max(d.A1$rugged), length.out = 30) 

mu_africa <- link(m7.1, data = data.frame(rugged = rugged_seq_africa))
mu.mean <- apply(mu_africa, 2, mean)
mu.PI <- apply(mu_africa, 2, HPDI, prob = 0.89) 


plot(log_gdp ~ rugged, data = d.A1, col = col.alpha(rangi2, 0.5))
mtext("Africa")
#draw MAP line
lines(rugged_seq_africa, mu.mean)

#draw HPDI region for line
shade(mu.PI, rugged_seq_africa)


```


```{r}
#define sequence of rugged terrain values to make predictions on
rugged_seq_not_africa <- 
  seq(from = min(d.A0$rugged), to = max(d.A0$rugged), length.out = 30) 

mu_not_africa <- link(m7.2, data = data.frame(rugged = rugged_seq_not_africa))
mu.mean <- apply(mu_not_africa, 2, mean)
mu.PI <- apply(mu_not_africa, 2, HPDI, prob = 0.89) 


plot(log_gdp ~ rugged, data = d.A0, col = col.alpha(rangi2, 0.5))
mtext("not Africa")
#draw MAP line
lines(rugged_seq_not_africa, mu.mean)

#draw HPDI region for line
shade(mu.PI, rugged_seq_not_africa)

```

Its not a good idea to split data, like we split the data above. The following are the reasons for it:

1) Some parameters, such as sigma, were assumed to not depend on a country's Africa identity. By splitting the data, we are hurting the accuracy of such estimates because we are now making two less accurate estimates instead of pooling all data into making one solid estimate. 

2) In order to make probability estimates about the `cont_africa` variable, we need to include it in the data that the model works on. Unless our model analyses all the data, it canot make a solid estimate of the uncertainty in the probability that `cont_africa` has on the outcome variable.

3) We might need to use information criteria to compare how different models treat different continents. In order for the estimate to be as accurate as possible, the models should read all the data.

4) In multilevel models, we can in fact learn across different categories. Therefore, it makes sense to not split data before inputting it into the model. 

Lets see if adding the categorical variable(dummy variable) `cont_africa` will reveal the reverse slope in the model.

R Code 7.3

The following is a simple linear regression model for `log_gdp` on ruggedness over the entire dataset  

```{r}
m7.3 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)
```

The second is a model that includes a dummy variable for African countries.

R Code 7.4

```{r}
m7.4 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged + bA * cont_africa,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)

```

Now, lets compare the WAIC values of these models

R Code 7.5

```{r}
compare(m7.3, m7.4)
```

From the above, we can see that all weight is assigned to model m7.4. The standard error in the difference of WAIC between the two models is 15, but the difference in WAIC values is 63, which implies a 95% confidence interval difference of 63 +- 30. Thus, we can gauge from these numbers that the dummy variable is picking up some important information since m7.4 is superior to the simple model. 

Now, lets plot the posterior prediction of m7.4 to see whether it achieves different slopes within and outside of Africa. 

R Code 7.6

```{r}
rugged.seq <- seq(from = -1, to = 8, by = 0.25)

#compute mu over samples, fixing cont_africa = 0
mu.NotAfrica <- 
  link(m7.4, data = data.frame(cont_africa = 0, rugged = rugged.seq))

#compute mu over samples, fixing cont_africa = 1
mu.Africa <- 
  link(m7.4, data = data.frame(cont_africa = 1, rugged = rugged.seq))

#summarise to means and intervals 
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob = 0.97)
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob = 0.97)

```

```{r}
plot(log_gdp ~ rugged, data = dd) 

#plot the regression lines
lines(rugged.seq, mu.NotAfrica.mean)
lines(rugged.seq, mu.Africa.mean, col = col.alpha(rangi2, 0.5))

#draw HPDI region for lines
shade(mu.NotAfrica.PI, rugged.seq)
shade(mu.Africa.PI, rugged.seq, col = col.alpha(rangi2, 0.5))
mtext("Africa is in Blue, Other Countries in Black")


```

From the above, we can see that by just adding a dummy variable to the model, the model still predicted a poor relationship between ruggedness and a country's GDP. What the model only did was that it predicted a lower average GDP values for African countries, which can be shown from the fact that the regression line for Africa is lower. 

Now, instead, lets add a linear model that models the linear interaction effect between ruggedness and African nations. In short, the relationship between GDP and ruggedness will be conditional on whether a country is in Africa or not. 

Yi ~ Normal(µi, σ) [likelihood] 
µi = α + γiRi + βAAi [linear model of µ]
γi = βR + βARAi [linear model of slope] 

The third line above is the linear interaction effect between ruggedness and whether a country is in Africa or not. 

Lets model the above linear interaction.

R Code 7.7

```{r}
m7.5 <- rethinking::map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + gamma * rugged + bA * cont_africa,
    gamma <- bR + bAR * cont_africa,
    a ~ dnorm(8, 100),
    bA ~ dnorm(0, 1),
    bR ~ dnorm(0, 1),
    bAR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = dd
)
```

Now, lets compare this model using WAIC values with the previous models.

```{r}
compare(m7.3, m7.4, m7.5)
```

Model 7.5 has a weight of 0.97. However, the weight of 0.03 that is given to model m7.4 shows that there is slight overfitting happening in m7.5. Also, the difference in the standard errors between the top two models is almost the same. This phenonema might also be there because there are only a limited number of African countries, and hence there is some sparsity in the data. 



