---
title: "Chapter 6"
author: "Usman Khaliq"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(rethinking)
# Parameters

#===============================================================================

# Code
```

R Code 6.1

Lets simulate data for 7 hominin species, of their average brain volumes and body mass

```{r}
sppnames <- c(
  "afarensis", 
  "africanus",
  "habilis",
  "boisei",
  "rudolfensis",
  "ergaster",
  "sapiens"
  ) 

brainvolcc <- c(438, 452, 612, 521, 752, 871, 1350)

masskg <- c(37.0, 35.5, 34.5, 41.5, 55.5, 61.0, 53.5)

d <- data.frame(species = sppnames, brain = brainvolcc, mass = masskg)
```

Lets fit this data with increasingly complex models.

First, lets fit it in a linear model that assumes tha the average brain volume of a species is a linear function of its body mass.

R Code 6.2

```{r}
m6.1 <- lm(brain ~ mass, data = d)
```

Now, lets compute the R2 of this model.

R Code 6.3

```{r}
1 - var(resid(m6.1)) / var(d$brain)
```

```{r}
summary(m6.1)
```

R Code 6.4

Now, lets use a more complex model that uses a second degree polynomial.

```{r}
m6.2 <- lm(brain ~ mass + I(mass^2), data = d)
m6.3 <- lm(brain ~ mass + I(mass^2) + I(mass^3), data = d)
m6.4 <- lm(brain ~ mass + I(mass^2) + I(mass^3) + I(mass^4), data = d)
m6.5 <- lm(brain ~ mass + I(mass^2) + I(mass^3) + I(mass^4) + I(mass^5), data = d)
m6.6 <- 
  lm(
    brain ~ mass +
      I(mass^2) +
      I(mass^3) +
      I(mass^4) +
      I(mass^5) +
      I(mass^6),
    data = d
  )

```

R Code 6.9

```{r}
p <- c(0.3, 0.7)
-sum(p * log(p))
```

R Code 6.10

Lets use `logLik` to compute deviance

```{r}
#fit model with lm
m6.1 <- lm(brain ~ mass, d)

#compute deviance by cheating
(-2) * logLik(m6.1)
```

